---
title: E1
subtitle: Alien species eradicated
author: Hanno Sandvik
date: 12 March 2023
output:
  md_document:
    toc: true
---

## Introduction
This code calculates indicator E1 of [NorInAliS](https://github.com/HannoSandvik/NorInAliS).


## Preparations
Load the establishment categories from the Alien Species List 2018:
```{r}
EstablishmentCategories2018  <- read.csv2("../cache/a-e.csv", as.is=T)
```
This file consists of the two columns "Name" and "CategN" of the sheet "List I" 
of the Excel workbook which is Online Resource 1 of 
[Sandvik et al. (2019)](https://doi.org/10.1007/s10530-019-02058-x):
```{r}
head(EstablishmentCategories2018)
```


Define auxiliary functions:
```{r}
# Combines text strings
"%+%" <- function(string1, string2) paste0(string1, string2)


# Calculates the union of two sets (vectors)
"%A%" <- function(set1, set2)
  if (is.null(set1)) logical(0) else as.vector(na.omit(set1[set1 %in% set2]))
```


## Functions
Define an indicator-specific function:
```{r}
E1 <- function(List1, List2,
               ID1 = "Name", ID2 = "Name",
               EstCat1 = "CategN", EstCat2 = "CategN") {
  # It is crucially important that the identifiers are identical in the two lists.
  # Here, the species names are used as identifiers. 
  # However, species names may change due to taxonomic revision.
  # Therefore, the use of unequivocal (e.g. numerical) species identifiers is
  # to be preferred. 
  w1 <- which(List1[, EstCat1] %in% c("C2", "C3", "D1", "D2", "E"))
    # species that are reproducing unaidedly at time 1
  w2 <- which(List2[, EstCat2] == "A")
    # species that are absent from the assessment area at time 2
  eradicated <- List1[w1, ID1] %A% List2[w2, ID2]
    # union of the two sets of identifiers
  return(length(eradicated))
}
```


## Calculations
Indicator E1 presupposes that establishment categories according to 
[Blackburn et al.'s (2011)](http://dx.doi.org/10.1016/j.tree.2011.03.023) 
"unified framework" are available for two consecutive Alien Species Lists. 
This information will be an integrative part of the coming Alien Species List
(2023, in prep.). For the current Alien Species List
([Artsdatabanken 2018](https://artsdatabanken.no/fremmedartslista2018)), 
establishment categories were published by 
[Sandvik et al. (2019)](https://doi.org/10.1007/s10530-019-02058-x).
For the previous Alien Species List 
([Artsdatabanken 2012](https://www.biodiversity.no/Pages/201621/)), 
this information is not available.
This means that, currently, indicator E1 cannot be estimated automatically.
A "manual" comparison between the Alien Species Lists 2012 and 2018 shows that
no alien has been eradicated between 2012 and 2018.
The following script shows how this test can be automatised,
albeit as an illustration only.


**Illustration of the function:**

Since the establishment categories for the Alien Species List 2012 are not
available, the actual result can be "simulated" as follows:
```{r}
EstablishmentCategories2012 <- EstablishmentCategories2018
cat("Actual estimate for E1: " %+%
    E1(EstablishmentCategories2012, EstablishmentCategories2018) %+% "\n")
```


**Further illustration:**

Just to illustrate the method, we may define a dummy dataset with shuffled
establishment categories:
```{r}
for (i in 1:9) {
  DummyCategories <- EstablishmentCategories2018
  DummyCategories[, "CategN"] <- sample(DummyCategories[, "CategN"])
  cat("Randomised estimate for E1 (randomisation " %+% i %+% "): " %+%
      E1(DummyCategories, EstablishmentCategories2018) %+% "\n")
}
```

