---
title: P1
subtitle: Number of alien species
author: Hanno Sandvik
date: 14 March 2023
output:
  md_document:
    toc: true
---

## Introduction
This code calculates indicator P1 of [NorInAliS](https://github.com/HannoSandvik/NorInAliS).


## Preparations
Load data from the Alien Species List 2018 
([Sandvik et al. 2020](https://doi.org/10.5061/dryad.8sf7m0cjc)):
```{r}
fab  <- read.csv2(url("https://datadryad.org/stash/downloads/file_stream/359484"),
                  as.is=T)
```


Restrict data to alien species that are reproducing unaidedly in mainland Norway:
```{r}
fab <- fab[which(fab$Status == "reproducing" & fab$Mainl),]
```


Define an auxiliary function:
```{r}
# Combines text strings
"%+%" <- function(string1, string2) paste0(string1, string2)
```


## Functions
Define indicator-specific functions:
```{r}
as.nr <- function(x)
  # assigns numerical values from 1 to 5 to the five ecological impact categories
  return(sapply(x, function(z) switch(z, "SE"=5,"HI"=4,"PH"=3,"LO"=2,"NK"=1,NA)))


P1a <- function(dataset, column = "Impact")
  # estimates P1(a)
  return(length(which(dataset[, column] %in% c("SE", "HI", "PH", "LO", "NK"))))
  

P1b <- function(dataset, column = c("minImp", "Impact", "maxImp")) {
  # estimates P1(b)
  q0 <-       sum(as.nr(dataset[, column[1]]))  # minimum and 2.5% confidence level
  q1 <- round(sum(as.nr(dataset[, column[2]]) * 0.75 +
                  as.nr(dataset[, column[1]]) * 0.25))  # 1st quartile
  q2 <-       sum(as.nr(dataset[, column[2]]))  # best estimate (mean and median)
  q3 <- round(sum(as.nr(dataset[, column[2]]) * 0.75 +
                  as.nr(dataset[, column[3]]) * 0.25))  # 3rd quartile
  q4 <-       sum(as.nr(dataset[, column[3]]))  # maximum and 97.5% confidence level
  SD <- round(sd(c(rep(q0,   floor(P1a(dataset, column[2]) / 4)),
                   rep(q2, ceiling(P1a(dataset, column[2]) / 2)),
                   rep(q4,   floor(P1a(dataset, column[2]) / 4)))))
                   # approximation of the S.D.
  p1b <- c(q2, SD, q0, q1, q2, q3, q4)
  names(p1b) <- c("Average", "St.Dev.", c(2.5, 25, 50, 75, 97.5) %+% "%CL")
  return(p1b)
}
```

Note that these functions presuppose that the impact categories follow the 
[GEIAA](https://doi.org/10.1007/s10530-019-02033-6) protocol. 
For other impact assessment schemes, the functions would have to be adjusted.


## Calculations
P1(a) is simply the number of reproducing alien species in mainland Norway
```{r}
cat("P1(a): " %+% P1a(fab) %+% "\n")
```


The disaggregation for ecological impact categories looks like this:
```{r}
{ 
  cat("Alien species in Norway by impact category:\n")
  print(table(fab$Impact)[c("NK", "LO", "PH", "HI", "SE")])
}
```


P1(b) is a weighted sum of impact categories. 
It takes the uncertainty in impact categories into account,
quantified as quartiles:
```{r}
{
  cat("P1(b):\n")
  print(P1b(fab))
}
```

