---
title: P5
subtitle: Alien species reported from monitoring
author: Hanno Sandvik
date: 22 March 2023
output:
  md_document:
    toc: true
---

## Introduction
This code calculates indicator P5 of [NorInAliS](https://github.com/HannoSandvik/NorInAliS).


## Preparations
Load the `sf` package ([Pebesma 2018](https://doi.org/10.32614/RJ-2018-009)):
```{r}
library(sf)
```


Download data from the "spatially representative nature monitoring program for terrestrial ecosystems", 
or **ANO**, by its Norwegian acronym 
([Norwegian Environment Agency 2022](https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054)):
```{r}
download.file("https://nedlasting.miljodirektoratet.no/naturovervaking/naturovervaking_eksport.gdb.zip",
              "ANO.gdb.zip")
```


Extract the layer containing ANO's survey points:
```{r}
{
  unzip("ANO.gdb.zip", exdir = "P5")
  ANO <- st_read("P5/Naturovervaking_eksport.gdb", layer = "ANO_SurveyPoint")
  unlink("P5", recursive = TRUE, force = TRUE)
  file.remove("ANO.gdb.zip")
}
```


Restrict data to information collected in 2019 and 2020:
```{r}
years <- 2019:2020
ANO <- ANO[which(ANO$aar %in% years), ]
```


Exclude inaccessible survey circles:
```{r}
ANO <- ANO[-which(ANO$utilgjengelig_punkt == "ja"), ]
```


Check the number of survey points in the selected portion of the ANO dataset:
```{r}
nrow(ANO)
```


Download data from the Alien Species List 2018 
([Sandvik et al. 2020)](https://doi.org/10.5061/dryad.8sf7m0cjc):
```{r}
fab  <- read.csv2(url("https://datadryad.org/stash/downloads/file_stream/359484"),
                  as.is=T)
```


Restrict these data to alien plants that are reproducing unaidedly in mainland Norway, 
and that have been risk-assessed to severe (SE), high (HI) or potentially high (PH) 
ecological impact:
```{r}
fab <- fab[which(fab$Kingd == "Plantae" & fab$Status == "reproducing" & 
                 fab$Mainl & fab$Impact %in% c("SE", "HI", "PH")),]
```


Define auxiliary functions:
```{r}
# Combines text strings
"%+%" <- function(string1, string2) paste0(string1, string2)


# I just find this more intuitive...
"%contains%" <- function (textstring, searchtext)
  grepl(searchtext, textstring, fixed = TRUE)


# Prepare species names for comparison
prepare.name <- function(x) {
  x <- gsub("_",      " ", x)
  x <- gsub(" × ",    " ", x)
  x <- gsub(" ×",     " ", x)
  x <- gsub(" x ",    " ", x)
  x <- gsub(" \xd7 ", " ", x)
  x <- gsub(" \xd7",  " ", x)
  return(tolower(x))
}
```


The ANO variable `hovedoekosystem_punkt` should contain the main ecosystem 
in each survey point:
```{r}
table(ANO$hovedoekosystem_punkt, useNA="always")
```


Translate the main ecosystems (MES) into English:
```{r}
ANO$MES <- NA
ANO$MES[which(ANO$hovedoekosystem_punkt == "arktisk_tundra")]  <- "Arctic"
ANO$MES[which(ANO$hovedoekosystem_punkt == "ferskvann")]       <- "freshwater"
ANO$MES[which(ANO$hovedoekosystem_punkt == "fjell")]           <- "alpine"
ANO$MES[which(ANO$hovedoekosystem_punkt == "hav")]             <- "marine"
ANO$MES[which(ANO$hovedoekosystem_punkt == "Nakent berg")]     <- "alpine"
ANO$MES[which(ANO$hovedoekosystem_punkt == "naturlig_apne")]   <- "open lowlands"
ANO$MES[which(ANO$hovedoekosystem_punkt == "semi_naturlig_mark")] <- 
                                                                  "semi-natural"
ANO$MES[which(ANO$hovedoekosystem_punkt == "skog")]            <- "woodlands"
ANO$MES[which(ANO$hovedoekosystem_punkt == "sno_og_issystemer")] <- 
                                                                  "snow/ice"
ANO$MES[which(ANO$hovedoekosystem_punkt == "Sterkt endret")]   <- "strongly altered"
ANO$MES[which(ANO$hovedoekosystem_punkt == "sterkt_endret_mark")] <- 
                                                                  "strongly altered"
ANO$MES[which(ANO$hovedoekosystem_punkt == "vaatmark")]        <- "wetlands"
ANO$MES[which(ANO$hovedoekosystem_punkt ==
              "Utilgjengelig punkt i bratt fjellvegg. ")]      <- "inaccessible"
table(ANO$MES, useNA="always")
```


For some reason, the information is missing for the majority of survey points. 
This seems to be a problem that applies to the first two years of ANO data. 
So we now describe code in two subsections, one with code that should work starting from 2021, 
and one for 2019 and 2020, which are used in the baseline estimates.


### Code for future use (2021 and onwards)
This code has been tested for ANO data from 2021 and should hopefully work 
for future years as well. 
Note that this code is here only _shown_ and not _run_!


Restrict the dataset to the relevant main ecosystems:
```{r, eval=FALSE, echo=TRUE}
ANO <- ANO[which(ANO$MES %in% c("alpine", "Arctic", "open lowlands",
                                "semi-natural", "wetlands", "woodlands")), ]
```


Check how many survey points are left after we have excluded the irrelevant ones:
```{r, eval=FALSE, echo=TRUE}
nrow(ANO)
```


Alien species are recorded in the variable `fa_registrert` 
(and two or more species are delimited using " + ").
This allows to count the number of alien species per ANO survey circle:
```{r, eval=FALSE, echo=TRUE}
AlienSpecies <- c()
ANO$number_of_alien_species <- 0
for (i in 1:nrow(ANO)) {
  if (!is.na(ANO$fa_registrert[i])) {
    spec <- unlist(strsplit(ANO$fa_registrert[i], " + ", fixed = TRUE))
    ANO$number_of_alien_species[i] <- length(spec)
    AlienSpecies <- c(AlienSpecies, spec)
  }
}
```


It might happen that an alien species has been recorded in the variable 
`art_alle_registrert`, but that it has not been listed in `fa_registrert`. 
To correct for such oversights, we need to compare the recorded names 
with the Alien Species List:
```{r, eval=FALSE, echo=TRUE}
ANOnames <- prepare.name(ANO$art_alle_registrert)
for (i in prepare.name(fab$Name)) {
  w <- which(ANOnames %contains% i)
  if (length(w)) {
    for (j in w) {
      if (!(i %in% prepare.name(ANO$fa_registrert[j]))) {
        if (is.na(ANO$fa_registrert[j])) {
          ANO$fa_registrert[j] <- i
        } else {
          ANO$fa_registrert[j] <- ANO$fa_registrert[j] %+% " + " %+% i
        }
        v <- which(prepare.name(AlienSpecies) %contains% i)
        if (length(v)) {
          i <- AlienSpecies[v[1]]
        }
        AlienSpecies <- c(AlienSpecies, i)
        ANO$number_of_alien_species[j] <- ANO$number_of_alien_species[j] + 1
      }
    }
  }
}
rm(ANOnames)
```


It may happen that an alien species _cover_ > 0 is reported without 
an alien species _name_, or vice versa. It is more likely that _missing_ values are 
an oversight than that _reported_ values are erroneous. That's at least prudent 
to assume. We therefore adjust the number of alien species according to the 
cover variable and vice versa (minimum estimates in both cases, i.e.
assuming 1 species where none has been reported, 
and a cover of 0.1% where none has been reported):
```{r, eval=FALSE, echo=TRUE}
ANO$number_of_alien_species[         which(ANO$fa_total_dekning > 0)] <-
  sapply(ANO$number_of_alien_species[which(ANO$fa_total_dekning > 0)], 
         max, 1, na.rm = TRUE)
ANO$fa_total_dekning[         which(ANO$number_of_alien_species > 0)] <-
  sapply(ANO$fa_total_dekning[which(ANO$number_of_alien_species > 0)], 
         max, 0.1, na.rm = TRUE)
```


### Code for the baseline estimate (2019 and 2020)
The code in this subsection is run.

We can try to fill in the missing main ecosystems using other ANO variables. 
Let's use the variable `kartleggingsenhet_250m2`, which contains the 
ecosystem mapping units for survey circles, following the "Nature in Norway" system 
(see [Halvorsen et al. 2020](http://dx.doi.org/10.1111/geb.13164)):
```{r}
ANO$MES[which(ANO$kartleggingsenhet_250m2 == "ferskvann")]     <- "freshwater"
ANO$MES[which(ANO$kartleggingsenhet_250m2 == "hav")]           <- "marine"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2,1,1) == "F")] <- "freshwater"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2,1,1) == "H")] <- "marine"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2,1,1) == "I")] <- "ice/snow"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2,1,1) == "L")] <- "freshwater"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2,1,1) == "M")] <- "marine"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2,1,1) == "O")] <- "freshwater"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2,1,1) == "V")] <- "wetlands"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2 %+% "-", 1, 3) %in%
                c("L4-",  "M8-" ))]                            <- "wetlands"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2 %+% "-", 1, 3) %in%
                c("T2-",  "T6-",  "T8-" ))]                    <- "open lowlands"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2 %+% "-", 1, 4) %in%
                c("T11-", "T12-", "T15-", "T18-", "T20-",
                  "T21-", "T23-", "T24-", "T25-", "T29-"))]    <- "open lowlands"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2 %+% "-", 1, 3) %in%
                c("T1-",  "T3-",  "T5-",  "T7-" ))]            <- "alpine"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2 %+% "-", 1, 4) %in%
                c("T13-", "T14-", "T16-", "T17-", "T19-",
                  "T22-", "T26-", "T27-"))]                    <- "alpine"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2 %+% "-", 1, 3) %in%
                c("T4-",  "V2-",  "V8-" ))]                    <- "woodlands"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2 %+% "-", 1, 4) %in%
                c("T30-"))]                                    <- "woodlands"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2 %+% "-", 1, 3) %in%
                c("T9-" ))]                                    <- "Arctic"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2 %+% "-", 1, 4) %in%
                c("T10-", "T28-"))]                            <- "Arctic"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2 %+% "-", 1, 3) %in%
                c("V9-" ))]                                    <- "semi-natural"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2 %+% "-", 1, 4) %in%
                c("T31-", "T32-", "T33-", "T34-", "V10-"))]    <- "semi-natural"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2 %+% "-", 1, 4) %in%
                ("T" %+% (35:45) %+% "-"))]                    <- "strongly altered"
ANO$MES[which(substr(ANO$kartleggingsenhet_250m2 %+% "-", 1, 4) %in%
                c("V11-", "V12-", "V13-"))]                    <- "strongly altered"
table(ANO$MES, useNA="always")
```


Still, some missing values. We can get a few more from `kartleggingsenhet_1m2`, 
which contains the ecosystem mapping units for the (centres of) ANO survey points:
```{r}
ANO$MES[which(is.na(ANO$MES) &
              substr(ANO$kartleggingsenhet_1m2, 1, 1) == "F")] <- "freshwater"
ANO$MES[which(is.na(ANO$MES) & 
              substr(ANO$kartleggingsenhet_1m2, 1, 1) == "H")] <- "marine"
ANO$MES[which(is.na(ANO$MES) & 
              substr(ANO$kartleggingsenhet_1m2, 1, 1) == "I")] <- "ice/snow"
ANO$MES[which(is.na(ANO$MES) & 
              substr(ANO$kartleggingsenhet_1m2, 1, 1) == "L")] <- "freshwater"
ANO$MES[which(is.na(ANO$MES) & 
              substr(ANO$kartleggingsenhet_1m2, 1, 1) == "M")] <- "marine"
ANO$MES[which(is.na(ANO$MES) & 
              substr(ANO$kartleggingsenhet_1m2, 1, 1) == "O")] <- "freshwater"
ANO$MES[which(is.na(ANO$MES) & 
              substr(ANO$kartleggingsenhet_1m2, 1, 1) == "V")] <- "wetlands"
ANO$MES[which((is.na(ANO$MES) | ANO$MES == "freshwater") & 
              substr(ANO$kartleggingsenhet_1m2 %+% "-", 1, 3) %in%
                c("L4-" ))]                            <- "wetlands"
ANO$MES[which((is.na(ANO$MES) | ANO$MES == "marine") & 
              substr(ANO$kartleggingsenhet_1m2 %+% "-", 1, 3) %in%
                c("M8-" ))]                            <- "wetlands"
ANO$MES[which(is.na(ANO$MES) &
              substr(ANO$kartleggingsenhet_1m2 %+% "-", 1, 3) %in%
                c("T2-",  "T6-",  "T8-" ))]                    <- "open lowlands"
ANO$MES[which(is.na(ANO$MES) &
              substr(ANO$kartleggingsenhet_1m2 %+% "-", 1, 4) %in%
                c("T11-", "T12-", "T15-", "T18-", "T20-",
                  "T21-", "T23-", "T24-", "T25-", "T29-"))]    <- "open lowlands"
ANO$MES[which(is.na(ANO$MES) &
              substr(ANO$kartleggingsenhet_1m2 %+% "-", 1, 3) %in%
                c("T1-",  "T3-",  "T5-",  "T7-" ))]            <- "alpine"
ANO$MES[which(is.na(ANO$MES) &
              substr(ANO$kartleggingsenhet_1m2 %+% "-", 1, 4) %in%
                c("T13-", "T14-", "T16-", "T17-", "T19-",
                  "T22-", "T26-", "T27-"))]                    <- "alpine"
ANO$MES[which((is.na(ANO$MES) | ANO$MES == "wetlands") &
              substr(ANO$kartleggingsenhet_1m2 %+% "-", 1, 3) %in%
                c("T4-",  "V2-",  "V8-" ))]                    <- "woodlands"
ANO$MES[which(is.na(ANO$MES) &
              substr(ANO$kartleggingsenhet_1m2 %+% "-", 1, 4) %in%
                c("T30-"))]                                    <- "woodlands"
ANO$MES[which(is.na(ANO$MES) &
              substr(ANO$kartleggingsenhet_1m2 %+% "-", 1, 3) %in%
                c("T9-" ))]                                    <- "Arctic"
ANO$MES[which(is.na(ANO$MES) &
              substr(ANO$kartleggingsenhet_1m2 %+% "-", 1, 4) %in%
                c("T10-", "T28-"))]                            <- "Arctic"
ANO$MES[which((is.na(ANO$MES) | ANO$MES == "wetlands") &
              substr(ANO$kartleggingsenhet_1m2 %+% "-", 1, 3) %in%
                c("V9-" ))]                                    <- "semi-natural"
ANO$MES[which((is.na(ANO$MES) | ANO$MES == "wetlands") &
              substr(ANO$kartleggingsenhet_1m2 %+% "-", 1, 4) %in%
                c("T31-", "T32-", "T33-", "T34-", "V10-"))]    <- "semi-natural"
ANO$MES[which(is.na(ANO$MES) & 
              substr(ANO$kartleggingsenhet_1m2 %+% "-", 1, 4) %in%
                ("T" %+% (35:45) %+% "-"))]                    <- "strongly altered"
ANO$MES[which((is.na(ANO$MES) | ANO$MES == "wetlands") & 
              substr(ANO$kartleggingsenhet_1m2 %+% "-", 1, 4) %in%
                c("V11-", "V12-", "V13-"))]                    <- "strongly altered"
table(ANO$MES, useNA="always")
```


There are still 110 survey points left for which we were unable to find the main ecosystem. 
We have to exclude those, in addition to inaccessible points, main ecosystems where we do 
not expect vascular plants (freshwater, ice/snow, marine, rock) or were we are not 
interested in alien species (strongly altered nature, such as gardens).
```{r}
ANO <- ANO[which(ANO$MES %in% c("alpine", "Arctic", "open lowlands",
                                "semi-natural", "wetlands", "woodlands")), ]
```


Check how many survey points are left after we have excluded the irrelevant ones:
```{r}
nrow(ANO)
```
That's not too bad.


Check which alien species have been recorded:
```{r}
table(ANO$fa_registrert, useNA="always")
```


That's a let-down. The ANO variable `fa_registrert` has not been used in 2019 and 2020. 
As a workaround, we can compare the variable `art_alle_registrert`, 
which lists all species recorded in the centre survey point, 
with the species names in the Alien Species List 2018:
```{r}
ANO$number_of_alien_species <- 0
AlienSpecies <- c()
ANOnames <- prepare.name(ANO$art_alle_registrert)
for (i in prepare.name(fab$Name)) {
  w <- which(ANOnames %contains% i)
  AlienSpecies <- c(AlienSpecies, rep(i, length(w)))
  ANO$number_of_alien_species[w] <- ANO$number_of_alien_species[w] + 1
  if (length(w)) {
    ANO$fa_registrert[w] <- ifelse(is.na(ANO$fa_registrert[w]),             i,
                                         ANO$fa_registrert[w] %+% " + " %+% i)
  }
}
rm(ANOnames)
```


Furthermore, alien species may have been mentioned in free text in the variable 
`kommentar_250m2_flate`, by their scientific or Norwegian vernacular name. 
This is how this variable looks like for the relevant ANO circles:
```{r}
unique(ANO$kommentar_250m2_flate[which(ANO$fa_total_dekning > 0)])
```


We extract the relevant species names:
```{r}
for (i in 1:7) {
  spec1 <- c("platanlønn", "pinus mugo", "rødhyll",
             "bergfuru", "mispel")[i]
  spec2 <- c("acer pseudoplatanus", "pinus mugo", "sambucus racemosa",
             "pinus uncinata", "cotoneaster spec.")[i]
  w <- which(prepare.name(ANO$kommentar_250m2_flate) %contains% spec1)
  if (length(w)) {
    for (j in w) {
      if (is.na(ANO$fa_registrert[j])) {
        ANO$fa_registrert[j] <- spec2
        AlienSpecies <- c(AlienSpecies, spec2)
        ANO$number_of_alien_species[j] <- ANO$number_of_alien_species[j] + 1
      } else {
        if (!(ANO$fa_registrert[j] %contains% spec2)) {
          ANO$fa_registrert[j] <- ANO$fa_registrert[j] %+% " + " %+% spec2
          AlienSpecies <- c(AlienSpecies, spec2)
          ANO$number_of_alien_species[j] <- ANO$number_of_alien_species[j] + 1
        }
      }
    }
  }
}
```


In quite some cases, "Lutz _or_ Sitka spruce" were reported. These should be 
counted as records of a single species. We arbitrarily choose Sitka spruce:
```{r}
w <- which(prepare.name(ANO$kommentar_250m2_flate) %contains% "sitka" |
           prepare.name(ANO$kommentar_250m2_flate) %contains% "lutz")
if (length(w)) {
  for (j in w) {
    if (is.na(ANO$fa_registrert[j])) {
      ANO$fa_registrert[j] <- "picea sitchensis"
      AlienSpecies <- c(AlienSpecies, "picea sitchensis")
      ANO$number_of_alien_species[j] <- ANO$number_of_alien_species[j] + 1
    } else {
      if (!(ANO$fa_registrert[j] %contains% "picea sitchensis" |
            ANO$fa_registrert[j] %contains% "picea lutzii"  )) {
        ANO$fa_registrert[j] <- ANO$fa_registrert[j] %+% " + " %+%
                                "picea sitchensis"
        AlienSpecies <- c(AlienSpecies, "picea sitchensis")
        ANO$number_of_alien_species[j] <- ANO$number_of_alien_species[j] + 1
      }
    }
  }
}
```


Check the available ANO data on alien species:
```{r}
table(ANO$fa_total_dekning > 0, ANO$number_of_alien_species, useNA="always")
```


This table shows that there are 51 cases in which the reported alien species 
_cover_ is greater than 0 but where no alien species _name_ has been recorded, 
as well as one case in which an alien species name has been recorded but the 
reported alien species cover is 0. 
We therefore adjust the number of alien species according to the 
cover variable and vice versa:
```{r}
ANO$number_of_alien_species[         which(ANO$fa_total_dekning > 0)] <-
  sapply(ANO$number_of_alien_species[which(ANO$fa_total_dekning > 0)], 
         max, 1, na.rm = TRUE)
ANO$fa_total_dekning[         which(ANO$number_of_alien_species > 0)] <-
  sapply(ANO$fa_total_dekning[which(ANO$number_of_alien_species > 0)], 
         max, 0.1, na.rm = TRUE)
```


Upper-case first letters in species names:
```{r}
AlienSpecies <- toupper(substr(AlienSpecies, 1, 1)) %+% substr(AlienSpecies, 2, 72)
```


## Calculations
This section definitely works now (i.e. for the baseline estimate), 
but it should also work in the future (from 2021 onwards), given that 
the relevant ANO variables keep their current names, definitions and usage.

Show the number of alien species recorded per ANO survey point:
```{r}
table(ANO$number_of_alien_species, useNA="always")
```
These figures are not entirely reliable because not all alien species names 
seem to have been recorded.


The alien species recorded were (with the same reservations as above, 
and the additional one that some _P. lutzii_ have been recorded as _P. sitchensis_):
```{r}
table(AlienSpecies)
```


Now, let's calculate the baseline estimates of the three indicator definitions.

**P5(a)** - average number of alien species recorded in survey circles of the 
spatially representative nature monitoring program
(with the reservation that the underlying figures are not reliable, see above):
```{r}
x <- sum(ANO$number_of_alien_species, na.rm = TRUE)
N <- nrow(ANO)
cat("P5(a): " %+% round(x / N, 5) %+% " ± " %+%
    round(sqrt(x * (1 - x / N)) / N, 5) %+% " (mean ± standard deviation)\n")
```


**P5(b)** - proportion of survey circles of the spatially representative 
nature monitoring program in which at least one alien species is recorded:
```{r}
x <- length(which(ANO$fa_total_dekning > 0))
N <- nrow(ANO)
cat("P5(b): " %+% round(100 * x / N, 3) %+% "% ± " %+%
    round(100 * sqrt(x * (1 - x / N)) / N, 3) %+% "% (mean ± standard deviation)\n")
```


**P5(c)** - average cumulative coverage of alien species recorded in survey circles 
of the spatially representative nature monitoring program:
```{r}
cat("P5(c): " %+% 
    round(sum(ANO$fa_total_dekning, na.rm = TRUE) / nrow(ANO), 4) %+% "%\n")
```


Confidence levels for P5(b):
```{r}
conf <-  qbinom(c(0.025, 0.25, 0.5, 0.75, 0.975), N, x / N) / N
names(conf) <- (c(0.025, 0.25, 0.5, 0.75, 0.975) * 100) %+% "%CL"
print(conf)
```


Disaggregation of P5(b) for main ecosystems:
```{r}
{
  tab <- table(ANO$MES[ANO$fa_total_dekning > 0])
  cat("Number of ANO survey circles with alien species:\n")
  print(tab)
  cat("\n\nProportion, i.e. indicator definition P5(b) by ecosystem:\n")
  print(tab / table(ANO$MES)[names(tab)])
  cat("\n\nStandard deviations:\n")
  print(sqrt(tab * (1 - tab / table(ANO$MES)[names(tab)])) /
        table(ANO$MES)[names(tab)])
}
```

