---
title: S1
subtitle: Introduction rate of novel alien species
author: Hanno Sandvik
date: 18 May 2023
output:
  md_document:
    toc: true
---

## Introduction
This code calculates indicator S1 of [NorInAliS](https://github.com/HannoSandvik/NorInAliS).


## Preparations
Load data from the Alien Species List 2018:

* The "fab"  data are available from [Sandvik et al. (2020)](https://doi.org/10.5061/dryad.8sf7m0cjc).
* The "path" data are available from [Sandvik et al. (2022)](https://doi.org/10.5061/dryad.4b8gthtg7).

```{r}
fab  <- read.csv2(url("https://datadryad.org/stash/downloads/file_stream/359484"),
                  as.is=T)
path <- read.csv2(url("https://datadryad.org/stash/downloads/file_stream/1823673"),
                  as.is=T)
```


Restrict data to alien species that are reproducing unaidedly in mainland Norway:
```{r}
fab <- fab[which(fab$Status == "reproducing" & fab$Mainl),]
```


Translate `NA`s into "unknown":
```{r}
for (i in 5:7) {
  path[which(is.na(path[, i])), i] <- "unknown"
}
```


Count observations of novel alien species per year:
```{r}
novelObs <- 0
for (year in 1780:2017) novelObs[year] <- length(which(fab$Observ == year))
```


The respective figures from the Alien Species List 2012 are added manually, 
as the underlying data are available 
[online](http://databank.artsdatabanken.no/FremmedArt2012), 
but not as a machine-readable dataset:
```{r}
novelObs.by2012 <- 0
novelObs.by2012[1980:2017] <- 
  c( 5,  4,  2,  3,  1, 14,  8,  4,  5,  8,  # 1980s
     3, 14,  9, 15, 18, 18,  7,  7,  9, 12,  # 1990s
    12, 15,  9,  8, 13, 20, 14, 17, 13,  5,  # 2000s
     6,  1,  0,  0,  0,  0,  0,  0)          # 2010s
```


Define auxiliary functions
```{r}
# Combines text strings
"%+%" <- function(string1, string2) paste0(string1, string2)

# Calculates a running mean
running.mean <- function(x, n) {
  z <- y <- x
  if (n > 1) {
    for (i in 2:n) {
      z <- c(z[-1], NA)
      y <- y + z
    }
  }
  return(y / n)
}
```


## Calculations
Estimate average, standard deviation and confidence intervals:
```{r}
{
  cat("Average  2011-2015: " %+% mean(novelObs[2011:2015])  %+% "\n")
  cat("St. Dev. 2011-2015: " %+%   sd(novelObs[2011:2015])  %+% "\n")
  for (p in c(" 2.5", "25.0", "50.0", "75.0", "97.5")) {
    cat(p %+% "% CL 2011-2015: " %+% 
      qpois(as.numeric(p) / 100, mean(novelObs[2011:2015])) %+% "\n"
    )
  }
  cat("Average  2012-2016: " %+% mean(novelObs[2012:2016])  %+% "\n")
  cat("St. Dev. 2012-2016: " %+%   sd(novelObs[2012:2016])  %+% "\n")
  cat("Average  2013-2017: " %+% mean(novelObs[2013:2017])  %+% "\n")
  cat("St. Dev. 2013-2017: " %+%   sd(novelObs[2013:2017])  %+% "\n")
}
```


Disaggregation of S1 for main pathway categories:
```{r}
Table <- matrix(0, 6, 2,
  dimnames=list(
    c("release", "escape", "contaminant", "stowaway", "corridor", "unaided"),
    c("average", "st.dev.")
  )
)
for (pathway in rownames(Table)) {
  records <- 0
  for (year in 2011:2015) {
    records[year - 2010] <- length(which(fab$Observ == year &
      fab$Impact %in% c("NK","LO","PH","HI","SE") &
      fab$Name %in% path$Name[path$Introd & 
                              path$Time != "future" &
                              path$Cat == pathway]))
  }
  Table[pathway, ] <- c(mean(records), sd(records))
}
print(Table)
```


Plot a graph of the temporal development since 1800:
```{r, fig.cap="Novel introductions per year (grey) and running five-year mean (blue line) from 1800 to 2017", fig.width=6, fig.height=4.8}
# To produce Figure B-1, add:
# png("S1-Fig1.png", 1500, 1200, res=180)
par(mai=c(0.96, 0.96, 0.06, 0.06))
plot(1796:2017, novelObs[1796:2017], ty="n",
  xlim=c(1800, 2020), ylim=c(0, 24), xaxs="i", yaxs="i",
  xlab="Year", ylab="Novel introductions per year", cex.axis=1.2, cex.lab=1.8
)
polygon(c(1800, 1800:2017, 2017), c(0, novelObs[1800:2017], 0), 
        col=grey(0.84), border=NA)
lines(1796:2017+2, running.mean(novelObs, 5)[1796:2017], lwd=4.8, col="blue")
axis(1, seq(1800, 2020, 10), F, T)
axis(2, 0:24, F, T)
box()
```


Plot a graph for the years 1988 to 2017 comparing the data from 2012 and 2018:
```{r, fig.cap="Novel introductions per year (grey) and running five-year mean (blue) from 1990 to 2017, according to the Alien Species Lists of 2012 and 2018, respectively", fig.width=7.2, fig.height=4.8}
# To produce Figure B-2, add:
# png("S1-Fig2.png", 1500, 1200, res=180)
par(mai=c(0.96, 0.96, 0.06, 0.06))
plot(1800:2017, novelObs[1800:2017], ty="l", col=grey(0.72), lwd=2.4,
  xlim=c(1988, 2019), ylim=c(0, 24), xaxs="i", yaxs="i",
  xlab="Year", ylab="Novel introductions per year", cex.axis=1.2, cex.lab=1.8
)
axis(1, 1990:2017, F, T)
axis(2,    0:24,   F, T)
lines(1980:2011,                novelObs.by2012[1980:2011],     col=grey(0.72), lwd=1.8, lty=2)
lines(1800:2017+2, running.mean(novelObs       [1800:2017], 5), col="blue",     lwd=4.8)
lines(1980:2011+2, running.mean(novelObs.by2012[1980:2011], 5), col="blue",     lwd=3.6, lty=2)
lines(2006:2009, rep(23,4), col=grey(0.72), lwd=2.4)
lines(2006:2009, rep(22,4), col=grey(0.72), lwd=1.8, lty=2)
lines(2006:2009, rep(21,4), col="blue",     lwd=4.8)
lines(2006:2009, rep(20,4), col="blue",     lwd=3.6, lty=2)
text(2009, 23, "yearly figures 2018",  pos=4, cex=1.2)
text(2009, 22, "yearly figures 2012",  pos=4, cex=1.2)
text(2009, 21, "five-year means 2018", pos=4, cex=1.2)
text(2009, 20, "five-year means 2012", pos=4, cex=1.2)
box()
```

