---
title: P2
subtitle: Extent of terrestrial alien species
author: Hanno Sandvik
date: 15 March 2023
output:
  html_document:
    toc: true
---

## Introduction
This code calculates indicator P2 of [NorInAliS](https://github.com/HannoSandvik/NorInAliS).


## Preparations
Load data from the Alien Species List 2018 
([Sandvik et al. 2020](https://doi.org/10.5061/dryad.8sf7m0cjc)):
```{r}
fab  <- read.csv2(url("https://datadryad.org/stash/downloads/file_stream/359484"),
                  as.is=T)
```
The AOOs (areas of occupancy) provided in the above dataset are only the _best_
estimates of the _total_ AOO for each species (i.e. the _median_ expert judgement
of the _real_ AOO, including "dark figures" or unreported occurrences). However,
in order to quantify uncertainty, P2 needs more than that, viz. the _low_ and 
_high_ estimates of the total AOO (i.e. lower and upper _quartiles_) as well as 
the _known_ AOO (i.e. excluding dark figures). These values are here read from a
separate file. Their source is https://artsdatabanken.no/fremmedartslista2018.
```{r}
aoo <- read.csv2("cache/aoo.txt", as.is=T)
```


Restrict data to alien species that are reproducing unaidedly in mainland Norway:
```{r}
fab <- fab[which(fab$Status == "reproducing" & fab$Mainl),]
```


Restrict the data to terrestrial species:
```{r}
w <- which(fab$LifeSt %in%
           c("lim", "lim,mar", "lim,par", "lim,mar,par", "mar", "mar,par"))
fab <- fab[-w,]
aoo <- aoo[-w,]
```


Make sure that the two data frames are compatible:
```{r}
if (all(fab$Name == aoo$Name)) {
  cat("Everything is fine.\n")
} else {
  cat("ERROR: For some reason, the two dataframes are not compatible!\n")
}
```


Define an auxiliary function:
```{r}
# Combines text strings
"%+%" <- function(string1, string2) paste0(string1, string2)
```


## Functions
Define indicator-specific functions:

The first two functions use maximum-likelihood estimation to infer the 
standard deviation of the AOOs, based on the best estimate (median), low estimate
(1st quartile) and high estimate (3rd quartile), 
and assuming a log-normal distribution.
```{r}
f <- function(s, mean, q1, q3) return(
  (q1 - qlnorm(0.25, log(mean) - exp(2*s)/2, exp(s)))^2 +
  (q3 - qlnorm(0.75, log(mean) - exp(2*s)/2, exp(s)))^2
)


findSD <- function(Ex, q1, q3) return(
  exp(optimise(f, c(-12, 12), mean=Ex, q1=q1, q3=q3)$min)
)


P2 <- function(dataset,
               column = c("known", "low", "best", "high"),
               nsim = 100000, 
               maxArea = 323800) {
  # simulates AOOs for all species, which is the basis of indicator P2
  N <- nsim     # random numbers per species
  M <- maxArea  # maximum possible area in km^2
  aoo <- dataset[, column]
  AOO <- matrix(0, N, nrow(aoo))
  for (i in 1:nrow(aoo)) {
    if (any(is.na(aoo[i, c("low", "best", "high")]))) {
      if (is.na(aoo$best[i])) {
        if (is.na(aoo$known[i])) {
          # if no AOO is provided, assume it is 0
          AOO[, i] <- 0
        } else {
          # of no total AOOs are provided, assume they equal the known AOO
          AOO[, i] <- aoo$known[i]
        }
      } else {
        # if no low and high estimates are provided, use the best estimate
        AOO[, i] <- aoo$best[i]
      }
    } else {
      if (aoo$best[i] == 0) {
        # if the best estimate is 0, keep it 0
        AOO[, i] <- 0
      } else {
        # if low, best and high estimates are provided, estimate the standard
        # deviation and generate log-normally distributed random numbers
        SD <- findSD(aoo$best[i], aoo$low[i], aoo$high[i])
        AOO[, i] <- qlnorm(runif(N), log(aoo$best[i]) - SD * SD / 2, SD)
        AOO[, i] <-  sapply(AOO[, i], min, M)  # constrain to area of Norway
        AOO[, i] <- ceiling(AOO[, i] / 4) * 4  # ensure multiples  of 4 km^2
      }
    }
  } # i 
  return(AOO)
} # P2
```


## Error checking and correction
Check for obvious errors in the original data:

**(1)** Is any low estimate greater than the corresponding best estimate? 
```{r}
w <- which(aoo$low > aoo$best)
if (length(w)) {
  print(aoo[w,])
}
```
This looks very much like a punching error in line 28. Most likely, the best
estimate should have been in the middle between the low and high estimate
(implying a dark figure of 1000 rather than 10). The error is corrected manually:
```{r}
aoo$best[w] <- 80000
```


**(2)** Is any high estimate less than the corresponding best estimate?
```{r}
w <- which(aoo$high < aoo$best)
if (length(w)) {
  print(aoo[w,])
}
```
Here, it seems that the low and high estimate simply haven't been provided,
and that, as a default, the known AOO was listed instead. In the absence of
other information, we have to assume the low and high estimates to be equal
to the best estimate:
```{r}
aoo$low[w] <- aoo$high[w] <- aoo$best[w]
```


**(3)** By definition, AOOs are multiples of 4 square kilometres. Some figures 
are incompatible with this definition. We solve this by rounding upwards:
```{r}
aoo[, 2:5] <- ceiling(aoo[, 2:5] / 4) * 4
```


**(4)** Is any AOO greater than the area of mainland Norway?
```{r}
w <- which(aoo$high > 323800)
if (length(w)) {
  print(aoo[w,])
}
```
Indeed. That is clearly an overestimate. The high estimate is reduced to
(a bit less than) the area of Norway. All other estimates are possible 
in principle, and are thus left unchanged.
```{r}
aoo$high[w] <- 323000
```


Show a graph of the distribution of the best estimates of total AOOs:
```{r}
par(mai=c(0.96, 0.96, 0.06, 0.06), lwd=1.8)
H <- hist(log10(as.numeric(aoo$best)), breaks=seq(0, 6, 1/3),
  xlab="Area of occupancy (km²)", ylab="Number of species", main="",
  xaxt="n", cex.axis=1.2, cex.lab=1.8, lwd=1.8, col=grey(0.84))
axis(1, 0:6, c(1, 10, 100, expression(10^3), expression(10^4), expression(10^5),
  expression(10^6)), cex.axis=1.2, lwd=1.8)
```


## Simulations
If you want P2 for _high- and severe-impact_ species only, do this first (not run):
```{r, eval=FALSE, echo=TRUE}
aoo <- aoo[which(fab$Impact %in% c("HI", "SE")),]
```
If you want P2 for the _remaining_ species only, do this first (not run):
```{r, eval=FALSE, echo=TRUE}
aoo <- aoo[which(fab$Impact %in% c("NK", "LO", "PH")),]
```
If you want P2 for _all_ alien species, do none of the above.

Then start the simulation:
```{r, eval=FALSE, echo=TRUE}
AOO <- P2(aoo)
avg   <-     mean(apply(AOO, 1, sum))
stdev <-       sd(apply(AOO, 1, sum))
conf  <- quantile(apply(AOO, 1, sum), c(0.025, 0.25, 0.5, 0.75, 0.975))
```

```{r, eval=TRUE, echo=FALSE}
# To save time during rendering, a previous run is loaded:
load("cache/P2.RData")
# This file contains the three variables `avg`, `stdev` and `conf`
```


Output of the results:
```{r}
{ 
  cat("P2 is " %+% round(avg,   -2) %+% " ± " %+%
                   round(stdev, -2) %+% " (mean ± SD)\n\n")
  cat("Confidence levels:\n")
  print(round(conf, -2))
}
```

