---
title: P3
subtitle: Extent of alien freshwater species
author: Hanno Sandvik
date: 16 March 2023
output:
  html_document:
    toc: true
---

## Introduction
This code calculates indicator P3 of [NorInAliS](https://github.com/HannoSandvik/NorInAliS).


## Preparations
Data from the reporting system for the Water Framework Directive (WFD) in Norway 
are not available in a machine-readable format but have to be downloaded manually 
(from [Vann-nett](https://vann-nett.no/portal/#/reportgenerator/234/Vannforekomster%20med%20p%C3%A5virkninger,%20p%C3%A5virkningsgrad,%20p%C3%A5virkningsgruppe,%20driver,%20effekt)). 
The data used here consist of the information on alien species in Norwegian 
waterbodies that was downloaded in 2020. 
The columns of this dataset are the unique WFD identifyer of the waterbody, 
the name of the waterbody, its category ("Lake", "River" or "Coastal"), 
the area (in square kilometres) of lake and coastal waterbodies, 
the length (in kilometres) of river waterbodies, the scientific name of the 
alien species recorded, its ecological impact category, and whether the species 
is regionally alien (i.e. displaced within Norway, `TRUE`) or 
nationally alien (`FALSE`).
```{r}
WFD <- read.csv("cache/WFD.csv", sep=";", dec=".", as.is=TRUE)
head(WFD)
```


Restrict the data to lake and river waterbodies:
```{r}
WFD <- WFD[which(WFD$Category %in% c("Lake", "River")), ]
```


Specify the total number and size of waterbodies 
(from [Vann-nett](https://vann-nett.no/portal/#/area/1/all)):
```{r}
Nlakes  <-   6803  # number of lake waterbodies
Alakes  <-  12040  # total area of lake waterbodies in square kilometres
Nrivers <-  23294  # number of river waterbodies
Lrivers <- 482328  # total length of river waterbodies in kilometres
```


Load data from the Alien Species List 2018 
([Sandvik et al. 2020](https://doi.org/10.5061/dryad.8sf7m0cjc)):
```{r}
fab <- read.csv2(url("https://datadryad.org/stash/downloads/file_stream/359484"),
                 as.is=T)
aoo <- read.csv2("cache/aoo.txt", as.is=T)
```


Restrict data to alien species that are reproducing unaidedly in mainland Norway:
```{r}
fab <- fab[which(fab$Status == "reproducing" & fab$Mainl),]
```


Restrict the data to freshwater species:
```{r}
w <- which(fab$LifeSt %in% c("lim", "lim,mar", "lim,par", "lim,mar,par"))
fab <- fab[w,]
aoo <- aoo[w,]
```


Define an auxiliary function:
```{r}
# Combines text strings
"%+%" <- function(string1, string2) paste0(string1, string2)
```


## Functions
There are no indicator-specific functions for P3.
However, for the estimation of indicator definition P3(e), the same functions as 
for indicator [P2 (see there)](P2.rmd) are needed (loaded invisibly).
```{r, eval=TRUE, echo=FALSE}
f <- function(s, mean, q1, q3) return(
  (q1 - qlnorm(0.25, log(mean) - exp(2*s)/2, exp(s)))^2 +
  (q3 - qlnorm(0.75, log(mean) - exp(2*s)/2, exp(s)))^2
)


findSD <- function(Ex, q1, q3) return(
  exp(optimise(f, c(-12, 12), mean=Ex, q1=q1, q3=q3)$min)
)


P2 <- function(dataset,
               column = c("known", "low", "best", "high"),
               nsim = 100000, 
               maxArea = 323800) {
  # simulates AOOs for all species, which is the basis of indicator P2
  N <- nsim     # random numbers per species
  M <- maxArea  # maximum possible area in km^2
  aoo <- dataset[, column]
  AOO <- matrix(0, N, nrow(aoo))
  for (i in 1:nrow(aoo)) {
    if (any(is.na(aoo[i, c("low", "best", "high")]))) {
      if (is.na(aoo$best[i])) {
        if (is.na(aoo$known[i])) {
          # if no AOO is provided, assume it is 0
          AOO[, i] <- 0
        } else {
          # of no total AOOs are provided, assume they equal the known AOO
          AOO[, i] <- aoo$known[i]
        }
      } else {
        # if no low and high estimates are provided, use the best estimate
        AOO[, i] <- aoo$best[i]
      }
    } else {
      if (aoo$best[i] == 0) {
        # if the best estimate is 0, keep it 0
        AOO[, i] <- 0
      } else {
        # if low, best and high estimates are provided, estimate the standard
        # deviation and generate log-normally distributed random numbers
        SD <- findSD(aoo$best[i], aoo$low[i], aoo$high[i])
        AOO[, i] <- qlnorm(runif(N), log(aoo$best[i]) - SD * SD / 2, SD)
        AOO[, i] <-  sapply(AOO[, i], min, M)  # constrain to area of Norway
        AOO[, i] <- ceiling(AOO[, i] / 4) * 4  # ensure multiples  of 4 km^2
      }
    }
  } # i 
  return(AOO)
} # P2
```


## Calculations
If you want P3 for _high- and severe-impact_ species only, do this first (not run):
```{r, eval=FALSE, echo=TRUE}
WFD <- WFD[which(WFD$Impact %in% c("HI", "SE")),]
aoo <- aoo[which(fab$Impact %in% c("HI", "SE")),]
```
If you want P3 for the _remaining_ species only, do this first (not run):
```{r, eval=FALSE, echo=TRUE}
WFD <- WFD[which(WFD$Impact %in% c("NK", "LO", "PH")),]
aoo <- aoo[which(fab$Impact %in% c("NK", "LO", "PH")),]
```
If you want P3 for _all_ alien species (**default**), do none of the above.


Create an empty matrix for the results:
```{r}
P3 <- matrix(0, 4, 7, dimnames = list(
  c("Lakes", "Rivers", "Total", "St.Dev."),
  c("P3(a)", "P3(b)", "P3(c)", "Area", "Length", "P3(d)", "P3(e)")
))
```


Estimate P3(a):
```{r}
P3[1, 1] <- length(unique(WFD$ID[which(WFD$Category == "Lake")]))
P3[2, 1] <- length(unique(WFD$ID[which(WFD$Category == "River")]))
P3[3, 1] <- length(unique(WFD$ID))
P3[4, 1] <- NA
```


Estimate P3(b):
```{r}
P3[1, 2] <- P3[1, 1] / Nlakes
P3[2, 2] <- P3[2, 1] / Nrivers
P3[3, 2] <- P3[3, 1] / (Nlakes + Nrivers)
P3[4, 2] <- sqrt(P3[3, 1] * (1 - P3[3, 1] / (Nlakes + Nrivers))) /
            (Nlakes + Nrivers)
```


Estimate P3(c):
```{r}
P3[1, 3] <- length(unique(WFD$ID[which(!WFD$Regional & WFD$Category == "Lake")]))
P3[2, 3] <- length(unique(WFD$ID[which(!WFD$Regional & WFD$Category == "River")]))
P3[3, 3] <- length(unique(WFD$ID[which(!WFD$Regional)]))
P3[4, 3] <- sqrt(P3[3, 3] * (1 - P3[3, 3] / (Nlakes + Nrivers))) /
            (Nlakes + Nrivers)
P3[1, 3] <- P3[1, 3] / Nlakes
P3[2, 3] <- P3[2, 3] / Nrivers
P3[3, 3] <- P3[3, 3] / (Nlakes + Nrivers)
```


Estimate P3(d):
```{r}
for (i in unique(WFD$ID[which(WFD$Category == "Lake")])) {
  A <- WFD$Area[which(WFD$ID == i)[1]]
  # to ensure that every waterbody is only counted once
  if (!is.na(A)) {
    P3[1, 4] <- P3[1, 4] + A
  }
}
for (i in unique(WFD$ID[which(WFD$Category == "River")])) {
  l <- WFD$Length[which(WFD$ID == i)[1]]
  # to ensure that every waterbody is only counted once
  if (!is.na(l)) {
    P3[2, 5] <- P3[2, 5] + l
  }
}
P3[1, 6]   <- P3[1, 4] / Alakes
P3[2, 6]   <- P3[2, 5] / Lrivers
P3[3, 6]   <- mean(P3[1:2, 6])
P3[4, 6]   <- NA
P3[,  4:5] <- round(P3[, 4:5])
P3[2:4, 4] <- NA
P3[ -2, 5] <- NA
```


Estimate P3(e):
```{r}
AOO <- P2(aoo)
avg   <-     mean(apply(AOO, 1, sum))
stdev <-       sd(apply(AOO, 1, sum))
conf  <- quantile(apply(AOO, 1, sum), c(0.025, 0.25, 0.5, 0.75, 0.975))
P3[1:2, 7] <- NA
P3[3,   7] <- round(avg)
P3[4,   7] <- round(stdev)
```


Output of the results, where

* **P3(a)** is the _number_ of river and lake waterbodies in which at least one alien species is recorded as an ongoing impact according to the Water Framework Directive,
* **P3(b)** is the _proportion_ of river and lake waterbodies in which at least one alien species is recorded as an ongoing impact according to the Water Framework Directive,
* **P3(c)** is the proportion of river and lake waterbodies in which at least one _nationally_ alien species is recorded as an ongoing impact according to the Water Framework Directive,
* **Area** is the cumulative _area_ of lake waterbodies (in square kilometres) in which at least one alien species is recorded as an ongoing impact according to the Water Framework Directive,
* **Length** is the cumulative _length_ of river waterbodies (in kilometres) in which at least one alien species is recorded as an ongoing impact according to the Water Framework Directive,
* **P3(d)** is the _proportion_ of the _area_ of lake waterbodies and of the _length_ of river waterbodies in which at least one alien species is recorded as an ongoing impact according to the Water Framework Directive,
* **P3(e)** is the _sum of areas of occupancies_ (in square kilometres) of all alien freshwater species that are recorded as reproducing unaidedly in Norway, including estimated dark figures of the areas of occupancy:
```{r}
print(P3)
```
Indicator definition P3(b) is the recommended one.

For indicator definitions P3(b), P3(c) and P3(e), 
it is possible to estimate confidence levels:
```{r}
q <- c(0.025, 0.25, 0.5, 0.75, 0.975)
CL <- matrix(0, 3, 5, dimnames=list(
  c("P3(b)", "P3(c)", "P3(e)"),
  names(conf)
))
CL[1, ] <- qbinom(q, Nlakes + Nrivers, P3[3, 2]) / (Nlakes + Nrivers)
CL[2, ] <- qbinom(q, Nlakes + Nrivers, P3[3, 3]) / (Nlakes + Nrivers)
CL[3, ] <- conf
print(CL)
```


Have a look at lake areas:
```{r}
{
  cat("Average area of lakes _with_  alien species: " %+%
      round(          P3[1, 4]  /           P3[1, 1],  2) %+% " km²\n")
  cat("Average area of lakes without alien species: " %+%
      round((Alakes - P3[1, 4]) / (Nlakes - P3[1, 1]), 2) %+% " km²\n")
}
```
Large lakes are overrepresented among lakes with alien species.

