---
title: P4
subtitle: Extent of marine alien species
author: Hanno Sandvik
date: 16 March 2023
output:
  html_document:
    toc: true
---

## Introduction
This code calculates indicator P4 of [NorInAliS](https://github.com/HannoSandvik/NorInAliS).


## Preparations
Define auxiliary functions:
```{r}
# Combines text strings
"%+%" <- function(string1, string2) paste0(string1, string2)


# I just find this more intuitive...
"%contains%" <- function (textstring, searchtext) grepl(searchtext, textstring)
```


Load data from the Alien Species List 2018 
([Sandvik et al. 2020](https://doi.org/10.5061/dryad.8sf7m0cjc)):
```{r}
fab  <- read.csv2(url("https://datadryad.org/stash/downloads/file_stream/359484"),
                  as.is=T)
```
The AOOs (areas of occupancy) provided in the above dataset are only the _best_
estimates of the _total_ AOO for each species (i.e. the _median_ expert judgement
of the _real_ AOO, including "dark figures" or unreported occurrences). However,
in order to quantify uncertainty, P2 needs more than that, viz. the _low_ and 
_high_ estimates of the total AOO (i.e. lower and upper _quartiles_) as well as 
the _known_ AOO (i.e. excluding dark figures). These values are here read from a
separate file. Their source is https://artsdatabanken.no/fremmedartslista2018.
```{r}
aoo <- read.csv2("cache/aoo.txt", as.is=T)
```


Restrict data to alien species that are reproducing unaidedly in mainland Norway:
```{r}
fab <- fab[which(fab$Status == "reproducing" & fab$Mainl),]
```


Restrict the data to marine species:
```{r}
w <- which(fab$LifeSt %in%
           c("mar", "mar,par"))
fab <- fab[w,]
aoo <- aoo[w,]
```


Restrict data to coastal and bottom-dwelling species (This is based on ecosystem 
codes defined by "Nature in Norway", according to the EcoSyst framework, see 
[Halvorsen et al. 2020](https://doi.org/10.1111/geb.13164)):
```{r}
w <- delete <- which(fab$Ecosys %contains% "H01" & !(fab$Ecosys %contains% "M"))
if (length(w)) {
  for (i in w) {
    if (fab$Ecosys[i] %contains% "H01-05") {
      if (!(paste(unlist(strsplit(fab$Ecosys[i], ",")) %-% "H01-05",
                  collapse=",") %contains% "H01")) {
        delete <- w %-% i
      }
    }
  }
  if (length(delete)) {
    cat(length(delete) %+% " marine species " %+% ifelse(length(delete) == 1,
        "is omitted because it is", "are omitted because they are") %+%
        " not coastal or bottom-dwelling:\n")
    for (i in delete) {
      cat("* " %+% fab$Name[i] %+% "\n")
    }
    fab <- fab[-delete,]
    aoo <- aoo[-delete,]
  }
}
```


Make sure that the two data frames are compatible:
```{r}
if (all(fab$Name == aoo$Name)) {
  cat("Everything is fine.\n")
} else {
  cat("ERROR: For some reason, the two dataframes are not compatible!\n")
}
```


Load the relevant data from the reporting system for the Water Framework Directive 
(WFD) in Norway (for details, [see indicator P3](P3.rmd)):
```{r}
WFD <- read.csv("cache/WFD.csv", sep=";", dec=".", as.is=TRUE)
```


Restrict the data to coastal waterbodies:
```{r}
WFD <- WFD[which(WFD$Category == "Coastal"), ]
head(WFD)
```


How many alien species are reported in the two datasets?
```{r}
{
  cat("Alien Species List 2018:   " %+% length(unique(fab$Name))    %+% " spp.\n")
  cat("Water Framework Directive: " %+% length(unique(WFD$Species)) %+% " spp.\n")
}
```


Specify the total number and size of waterbodies 
(from [Vann-nett](https://vann-nett.no/portal/#/area/1/all)):
```{r}
Ncoast <-  2283  # number of coastal waterbodies
Acoast <- 93649  # total area of coastal waterbodies in square kilometres
```


## Functions
The estimation of indicator definition P4(a) uses the same functions as 
indicator [P2 (see there)](P2.rmd). They are here loaded invisibly.
```{r, eval=TRUE, echo=FALSE}
f <- function(s, mean, q1, q3) return(
  (q1 - qlnorm(0.25, log(mean) - exp(2*s)/2, exp(s)))^2 +
  (q3 - qlnorm(0.75, log(mean) - exp(2*s)/2, exp(s)))^2
)


findSD <- function(Ex, q1, q3) return(
  exp(optimise(f, c(-12, 12), mean=Ex, q1=q1, q3=q3)$min)
)


P2 <- function(dataset,
               column = c("known", "low", "best", "high"),
               nsim = 100000, 
               maxArea = 323800) {
  # simulates AOOs for all species, which is the basis of indicator P2
  N <- nsim     # random numbers per species
  M <- maxArea  # maximum possible area in km^2
  aoo <- dataset[, column]
  AOO <- matrix(0, N, nrow(aoo))
  for (i in 1:nrow(aoo)) {
    if (any(is.na(aoo[i, c("low", "best", "high")]))) {
      if (is.na(aoo$best[i])) {
        if (is.na(aoo$known[i])) {
          # if no AOO is provided, assume it is 0
          AOO[, i] <- 0
        } else {
          # of no total AOOs are provided, assume they equal the known AOO
          AOO[, i] <- aoo$known[i]
        }
      } else {
        # if no low and high estimates are provided, use the best estimate
        AOO[, i] <- aoo$best[i]
      }
    } else {
      if (aoo$best[i] == 0) {
        # if the best estimate is 0, keep it 0
        AOO[, i] <- 0
      } else {
        # if low, best and high estimates are provided, estimate the standard
        # deviation and generate log-normally distributed random numbers
        SD <- findSD(aoo$best[i], aoo$low[i], aoo$high[i])
        AOO[, i] <- qlnorm(runif(N), log(aoo$best[i]) - SD * SD / 2, SD)
        AOO[, i] <-  sapply(AOO[, i], min, M)  # constrain to area of Norway
        AOO[, i] <- ceiling(AOO[, i] / 4) * 4  # ensure multiples  of 4 km^2
      }
    }
  } # i 
  return(AOO)
} # P2
```


## Error checking and correction
Check for obvious errors in the original data:

**(1)** Is any low estimate greater than the corresponding best estimate? 
```{r}
w <- which(aoo$low > aoo$best)
if (length(w)) {
  print(aoo[w,])
} else {
  cat("Everything is fine.\n")
}
```


**(2)** Is any high estimate less than the corresponding best estimate?
```{r}
w <- which(aoo$high < aoo$best)
if (length(w)) {
  print(aoo[w,])
} else {
  cat("Everything is fine.\n")
}
```


**(3)** By definition, AOOs are multiples of 4 square kilometres. Some figures 
are incompatible with this definition. We solve this by rounding upwards:
```{r}
aoo[, 2:5] <- ceiling(aoo[, 2:5] / 4) * 4
```


**(4)** Is any AOO greater than the area of mainland Norway?
```{r}
w <- which(aoo$high > 323800)
if (length(w)) {
  print(aoo[w,])
} else {
  cat("Everything is fine.\n")
}
```


## Simulations
If you want P2 for _high- and severe-impact_ species only, do this first (not run):
```{r, eval=FALSE, echo=TRUE}
aoo <- aoo[which(fab$Impact %in% c("HI", "SE")),]
WFD <- WFD[which(WFD$Impact %in% c("HI", "SE")),]
```
If you want P2 for the _remaining_ species only, do this first (not run):
```{r, eval=FALSE, echo=TRUE}
aoo <- aoo[which(fab$Impact %in% c("NK", "LO", "PH")),]
WFD <- WFD[which(WFD$Impact %in% c("NK", "LO", "PH")),]
```
If you want P2 for _all_ alien species (**default**), do none of the above.

Then start the simulation:
```{r}
AOO <- P2(aoo)
avg   <-     mean(apply(AOO, 1, sum))
stdev <-       sd(apply(AOO, 1, sum))
conf  <- quantile(apply(AOO, 1, sum), c(0.025, 0.25, 0.5, 0.75, 0.975))
```


Output of the results:
```{r}
{ 
  cat("P4(a) is " %+% round(avg,   -2) %+% " km² ± " %+%
                      round(stdev, -2) %+% " km² (mean ± SD)\n\n")
  cat("Confidence levels (in km²):\n")
  print(round(conf, -2))
}
```


## Calculation of WFD-based indicator definitions
Create a variable for the results:
```{r}
P4 <- matrix(round(avg), 1, 5, 
      dimnames=list("Coastal", c("P4(a)", "P4(b)", "P4(c)", "Area", "P4(d)")))
```


Estimate P4(b):
```{r}
P4[1, 2] <- length(unique(WFD$ID))
```


Estimate P4(c):
```{r}
P4[1, 3] <- P4[1, 2] / Ncoast
```


Estimate P4(d):
```{r}
P4[1, 4] <- 0
for (i in unique(WFD$ID)) {
  A <- WFD$Area[which(WFD$ID == i)[1]]
  # to ensure that every waterbody is only counted once
  if (!is.na(A)) {
    P4[1, 4] <- P4[1, 4] + A
  }
}
P4[1, 5] <- P4[1, 4] / Acoast
P4[1, 4] <- round(P4[1, 4])
```


Output of the results, where

* **P4(a)** is the _sum of areas of occupancies_ (in square kilometres) of all coastal or bottom-dwelling marine alien species that are recorded as reproducing unaidedly in Norway, including estimated dark figures of the areas of occupancy,
* **P4(b)** is the _number_ of coastal waterbodies in which at least one alien species is recorded as an ongoing impact according to the Water Framework Directive,
* **P4(c)** is the _proportion_ of coastal waterbodies in which at least one alien species is recorded as an ongoing impact according to the Water Framework Directive,
* **Area** is the cumulative _area_ of coastal waterbodies (in square kilometres) in which at least one alien species is recorded as an ongoing impact according to the Water Framework Directive,
* **P4(d)** is the _proportion_ of the _area_ of coastal waterbodies in which at least one alien species is recorded as an ongoing impact according to the Water Framework Directive:
```{r}
print(P4)
```

