---
title: S3
subtitle: Active pathways of secondary spread
author: Hanno Sandvik
date: 12 March 2023
output:
  html_document:
    toc: true
---

## Introduction
This code calculates indicator S3 of [NorInAliS](https://github.com/HannoSandvik/NorInAliS).


## Preparations
Load data from the Alien Species List 2018:

* The "fab"  data are available from [Sandvik et al. (2020)](https://doi.org/10.5061/dryad.8sf7m0cjc).
* The "path" data are available from [Sandvik et al. (2022)](https://doi.org/10.5061/dryad.4b8gthtg7).

```{r}
fab  <- read.csv2(url("https://datadryad.org/stash/downloads/file_stream/359484"),
                  as.is=T)
path <- read.csv2(url("https://datadryad.org/stash/downloads/file_stream/1823673"),
                  as.is=T)
```


Translate `NA`s into "unknown":
```{r}
for (i in 5:7) {
  path[which(is.na(path[, i])), i] <- "unknown"
}
```


Restrict data to alien species that are reproducing unaidedly in mainland Norway:
```{r}
fab <- fab[which(fab$Status == "reproducing" & fab$Mainl),]
```


Restrict data to current pathways of secondary spread:
```{r}
path <- path[!path$Introd & path$Time == "current" & path$Name %in% fab$Name, ]
```


Define an auxiliary function:
```{r}
# Combines text strings
"%+%" <- function(string1, string2) paste0(string1, string2)
```


## Functions
Define indicator-specific functions:
```{r}
asFreq  <- function(x) {
  # translates a frequency interval into a numerical value
  # Note that input is in events per decade,
  #  whereas output is in events per year!
  sapply(x, function(z) switch(z,
    "<1"=0.07, "1-8"=0.5, "9-19"=1.4, ">19"=4.6, NA))
}


asAbund <- function(x) {
  # translates an abundance interval into a numerical value
  sapply(x, function(z) switch(z,
    "1"=1, "2-10"=5, "11-100"=50, "101-1000"=500, ">1000"=5000, NA))
}


rincr <- function(n, min, max, r=TRUE) {
  # generates random numbers according to an increasing
  # triangular probability distribution
  min + sqrt(    if(r) runif(n) else 0.5) * (max - min)
}


rdecr <- function(n, min, max, r=TRUE) {
  # generates random numbers according to a decreasing
  # triangular probability distribution
  max - sqrt(1 - if(r) runif(n) else 0.5) * (max - min)
}


assignFrequencies <- function(n, pathways, pw, r=TRUE) {
  # assigns unknown frequencies according to the distribution of known frequencies,
  # separately for main pathway categories
  freq <- table(
    pathways$Freq[which(pathways$Freq != "unknown" & pathways$Cat == pw)]
  )[c("<1", "1-8", "9-19", ">19")]
  if (any(is.na(freq))) {
    freq[which(is.na(freq))] <- 0
  }
  freq <- freq / sum(freq)
  for (i in 4:2) {
    freq[i] <- sum(freq[1:i])
  }
  seq <- if (r) runif(n) else (1:n - 0.5) / n
  return(c("<1", "1-8", "9-19", ">19")[sapply(seq, function(x) which(freq > x)[1])])
}


S3a <- function(pathways) {
  # estimates S3(a) according to Equation ¤1
  return(length(unique(
    (pathways$Name %+% pathways$Subcat)[!pathways$Introd & pathways$Time == "current"]
  )))
}


S3b <- function(pathways, nsim=0, CL=c(0.025, 0.25, 0.5, 0.75, 0.975)) {
  # estimates S3(b) according to Equation ¤2
  pathways <- pathways[which(!pathways$Introd & pathways$Time == "current"),]
  results <- matrix(0, nrow(pathways), max(1, nsim))
  freq <- matrix(pathways$Freq, nrow(pathways), max(1, nsim))
  for (pw in unique(pathways$Cat)) {
    w <- which(pathways$Freq == "unknown" & pathways$Cat == pw)
    if (length(w)) {
      freq[w,] <- assignFrequencies(length(w) * max(1, nsim),
                                    pathways, pw, r = nsim >= 1)
    }
  }
  W <- freq == "<1"
  results[W] <-                rincr(sum(W), 0.01,  0.1, r = nsim >= 1)
  W <- freq == "1-8"
  results[W] <- if (nsim >= 1) runif(sum(W), 0.10,  0.9) else 0.5
  W <- freq == "9-19"
  results[W] <- if (nsim >= 1) runif(sum(W), 0.90,  1.9) else 1.4
  W <- freq == ">19"
  results[W] <-                rdecr(sum(W), 1.90, 10.0, r = nsim >= 1)
  if (nsim >=1 ) {
    freq <- c(mean(apply(results, 2, sum)),
              sd(apply(results, 2, sum)),
              quantile(apply(results, 2, sum), CL))
    names(freq) <- c("Average", "St.Dev.", (CL * 100) %+% "%CL")
    return(freq)
  } else {
    return(sum(asFreq(freq)))
  }
}


S3c <- function(pathways) {
  # estimates S3(c) according to Equation ¤3
  # (Note: this is a minimum implementation of indicator definition S3c!
  #  It illustrates the way S3c would be estimated, but it currently ignores
  #  uncertainty and simply assumes that the (utterly few) abundances reported
  #  are in fact representative of the unknown abundances, which is unlikely.)
  freq <- asFreq(pathways$Freq[!pathways$Introd & pathways$Time == "current"])
  w <- which(is.na(freq))
  if (length(w)) {
    freq[w] <- mean(freq[-w])
  }
  abund <- asAbund(pathways$Abund[!pathways$Introd & pathways$Time == "current"])
  w <- which(is.na(abund))
  if (length(w)) {
    abund[w] <- mean(abund[-w])
  }
  return(sum(freq * abund))
}
```


## Calculations
Summarise available data on frequency and abundance:
```{r}
{
  N <- length(which(!path$Introd & path$Time == "current"))
  cat("Distribution of frequencies of secondary spread events (per decade):\n")
  print(table(path$Freq )[c(1, 3, 4, 2, 5)] / N)
  cat("\n\nDistribution of abundances (individuals) per secondary spread event:\n")
  print(table(path$Abund)[c(4, 3, 2, 1, 5)] / N)
}
```


Show how frequencies of introduction vary between main pathway categories:
```{r}
summary(lm(asFreq(Freq) ~ Cat, data=path, subset=which(Freq != "unknown")))
```


Tabulate frequencies separately for each main pathway category:
```{r}
for (pw in unique(path$Cat)) {
  cat("\n\n" %+% pw %+% ":\n")
  print(table(path$Freq[which(path$Cat == pw)]))
}
```
This shows that "corridor" has only 1 occurrence with unknown frequencies. 
Therefore, corridors are excluded from S3b below.


Estimate the three indicator definitions for the current Alien Species List:
```{r, eval=FALSE, echo=TRUE}
S3b_results <- S3b(path[path$Cat != "corridor",], nsim=100000)
```

```{r, eval=TRUE, echo=FALSE}
# To save time during rendering, a previous run is loaded:
load("cache/S3.RData")
# This file contains a vector named `S3b_results`
```

``` {r}
{
  cat("\nIndicator value for S3(a): " %+%       S3a(path),    "\n")
  cat("\nIndicator value for S3(b):\n");  print(S3b_results)
  cat("\nIndicator value for S3(c): " %+% round(S3c(path)), "\n\n")
}
```


Disaggregate indicator S3(b) for main pathway categories:
```{r, eval=FALSE, echo=TRUE}
S3b_disaggr <- list()
for (pw in c("release", "contaminant", "stowaway", "unaided")) {
  cat("\n\n" %+% pw %+% ":\n")
  S3b_disaggr[[pw]] <- S3b(path[which(path$Cat == pw),], nsim=100000)
  print(S3b_disaggr[[pw]])
}
```

```{r, eval=TRUE, echo=FALSE}
# To save time during rendering, a previous run is loaded:
load("cache/S3.RData")
# This file contains a list named `S3b_disaggr`
for (pw in c("release", "contaminant", "stowaway", "unaided")) {
  cat("\n\n" %+% pw %+% ":\n")
  print(S3b_disaggr[[pw]])
}
```

